LEGWORK: A python package for computing the evolution and detectability \\ of stellar-origin gravitational-wave sources with space-based detectors

### 2021-11-16 src/ms.tex
missed path in vb figure
final updates to introduction

The potential to detect stellar-origin sources with LISA has been studied in the past \citep[e.g.][]{Nelemans+2001, Liu+2009, Liu+2014, Ruiter+2010, Belczynski+2010, Yu+2010, Nissanke+2012}. More recently, the direct detection of gravitational waves with ground-based detectors has led to renewed interest in this topic with many recent papers addressing the issue \citep[e.g.][]{Korol+2017, Korol+2018, Korol+2019, Korol+2020, Christian+2017, Kremer+2017, Kremer+2018, Lamberts+2018, Lamberts+2019, Fang+2019, Andrews+2020, Lau+2020, Breivik+2020, Breivik+2020a, Roebber+2020, Chen+2020, Sesana+2020, Shao+2021}.

@@

\lw{} is an open-source Python package designed to streamline the process of making predictions of LISA detection rates for stellar-origin binaries such that it is as fast, reliable and simple as possible. This goal makes \lw{} unique among other implementations of gravitational-wave tools in the literature, which focus on a more broad coverage of the gravitational-wave spectrum and source classes, rather than an optimised approach for certain sources. \citep[e.g.][]{Moore+2015,Yi2021}. With \lw{} one can evolve the orbits of a binary or a collection of binaries and calculate their strain amplitudes for any range of frequency harmonics. One can compute the sensitivity curve for LISA or other future gravitational wave detectors (e.g. TianQin's curve, or that of a custom instrument) and use it to compute the signal-to-noise ratio of a collection of sources. Furthermore, \lw{} provides tools to visualise all of the results with easy-to-use plotting functions. Finally, \lw{} is fully tested to check for consistency in the derivations described below.

@@

\lw{} has been developed with stellar-origin binary population studies in mind. Wagg et al.\ (in prep) investigates several populations of potential LISA sources (double black holes, black hole neutron stars and double neutron stars). They use \lw{} to evolve each synthesised source over the age of the Milky Way, make predictions about the LISA detectable population and explore how it varies with different binary physics assumptions. Additionally, Thiele et al.\ (in prep) examines the implications of assuming a metallicity-dependent binary fraction for the formation of close double white dwarfs (WDWDs) in the LISA frequency band. They use \lw{} to calculate the SNR of WDWDs and, in particular, apply a custom noise model that combines the LISA sensitivity curve from \citet{Robson+2019} with a new fit for the Galactic confusion noise based on their WDWD population.

@@

It is often convenient to convert between orbital frequency, $f_{\rm orb}$, and the semi-major axis, $a$, of a binary and this can be accomplished with Kepler's third law.
\begin{equation}{\lwFuncURL{utils}{get_a_from_f_orb}}
    a = \qty(\frac{G(m_1 + m_2)}{(2 \pi f_{\rm orb})^{2}})^{1/3},
    \label{eq:kepler3rd}
\end{equation}
where $G$ is the gravitational constant. Inversely we could write
\begin{equation}{\lwFuncURL{utils}{get_a_from_f_orb}}
    f_{\rm orb} = \frac{1}{2 \pi} \sqrt{\frac{G(m_1 + m_2)}{a^3}}.
    \label{eq:kepler3rd_reverse}
\end{equation}
For circular binaries, gravitational-wave emission occurs at twice the orbital frequency ($f_{\rm GW} = 2 f_{\rm orb}$). However, for eccentric binaries, we need to consider all frequency harmonics of gravitational-wave emission. These are defined such that the $n^{\rm th}$ harmonic frequency is
\begin{equation}{}
    f_{n} \equiv n \cdot f_{\rm orb}.
    \label{eq:nth_freq}
\end{equation}
It will be important to know the relative gravitational-wave power radiated into the $n^{\rm th}$ harmonic for a binary with eccentricity $e$ for the following strain and SNR calculations. This is given by \citet[][Eq.\,20]{Peters+1963}
\begin{equation}{\lwFuncURL{utils}{peters_g}}
    \begin{aligned}
        g(n, e) = \frac{n^{4}}{32} & \Big\{ \Big[ J_{n-2}(n e)-2 e J_{n-1}(n e)+\frac{2}{n} J_{n}(n e) \\
        &\left. +2 e J_{n+1}(n e)-J_{n+2}(n e)\Big]^{2}\right.\\
        &\left.\kern-\nulldelimiterspace\right.
            \!\!\begin{aligned}
            +\big(1-e^{2}\big)\Big[&J_{n-2}(n e)-2 J_{n}(n e) \\
            &+J_{n+2}(n e)\Big]^{2}
            \end{aligned}\\
        &+\frac{4}{3 n^{2}}\Big[J_{n}(n e)\Big]^{2}\Big\},
    \end{aligned}
    \label{eq:g(n,e)}
\end{equation}
where $J_{n}(v)$ is the Bessel function of the first kind. Thus, the \textit{sum} of $g(n, e)$ over all harmonics gives the factor by which the gravitational-wave emission is stronger for a binary of eccentricity $e$ over an equivalent circular binary. This enhancement factor is \citep[][Eq.\,17]{Peters+1963}
\begin{equation}{\lwFuncURL{utils}{peters_f}}
    F(e) \equiv \sum_{n = 1}^{\infty} g(n, e) = \frac{1 + \frac{73}{24} e^2 + \frac{37}{96} e^4}{(1 - e^2)^{7/2}}.
    \label{eq:eccentricity_enhancement_factor}
\end{equation}
A useful rule of thumb is that $F(0.5) \approx 5.0$, or in words, a binary with eccentricity $0.5$ loses energy to gravitational waves at 5 times the rate of an equivalent circular binary.

@@

\subsubsection{Eccentric binaries}
The orbital evolution is more complex for eccentric binaries since the semi-major axis and eccentricity both evolve simultaneously and depend on one another. The final expression cannot be solved analytically and require numerical integration. The semi major axis, $a$, and eccentricity, $e$, are related as \citep[][Eq.~5.11]{Peters+1964}
\begin{equation}{\lwFuncURL{utils}{get_a_from_ecc}}
    a(e) = c_0 \frac{e^{12/19}}{(1 - e^2)} \left(1 + \frac{121}{304} e^2\right)^{870/2299},
    \label{eq:a_from_e}
\end{equation}
where $c_0$ satisfies the initial conditions such that $a(e_0) = a_0$. The time derivative of the eccentricity, $e$, is given by
\begin{equation}{\lwFuncURL{evol}{de_dt}}
    \frac{\mathrm{d}e}{\mathrm{d}t} = -\frac{19}{12} \frac{\beta}{c_{0}^{4}} \frac{e^{-29 / 19}\left(1-e^{2}\right)^{3 / 2}}{\left[1+(121 / 304) e^{2}\right]^{\frac{1181}{2299}}},
    \label{eq:dedt}
\end{equation}
which we can integrate to find $e(t)$ and convert to $a(t)$ using Eq.~\ref{eq:a_from_e} (\docsLink{\lwFuncURL{evol}{evol_ecc}}).

@@

In addition, we implement the fit to Eq.~\ref{eq:t_merge_eccentric} from \citet{Mandel+2021} that approximates the merger time as
\begin{equation}{\lwFuncURL{evol}{t_merge_mandel_fit}}
    \begin{split}
        t_{\rm merge} \approx& t_{\rm merge, circ} (1 - e^2_0)^{7/2} \\
    &\times(1 + 0.27 e_0^{10} + 0.33 e^{20}_0 + 0.2 e_0^{1000} ),
    \end{split}
\end{equation}
which gives $t_{\rm merge}$ with an error below 3\% for eccentricities below $0.9999$. We additionally add a rudimentary polynomial fit to further reduce this error to below $0.5\%$. The user may specify whether to use this fit or perform the full integral when calculating merger times in \lw{}.

@@

\begin{figure}[htb]
    \centering
    \includegraphics[width=\columnwidth]{verification_binaries_on_sc.pdf}
    \caption{The LISA verification binaries from \citet{Kupfer+2018} plotted on the LISA sensitivity curve \citep{Robson+2019} for a 4-year LISA mission. Points are coloured by their primary mass.}
    \label{fig:verification_binaries_on_sc}
\end{figure}

>>>>

The potential to detect stellar-origin sources with LISA has been studied in the past by various groups \citep[e.g.][]{Nelemans+2001, Liu+2009, Liu+2014, Ruiter+2010, Belczynski+2010, Yu+2010, Nissanke+2012}. The recent direct detection of gravitational waves with ground-based detectors has led to renewed interest in this topic \citep[e.g.][]{Korol+2017, Korol+2018, Korol+2019, Korol+2020, Christian+2017, Kremer+2017, Kremer+2018, Lamberts+2018, Lamberts+2019, Fang+2019, Andrews+2020, Lau+2020, Breivik+2020, Breivik+2020a, Roebber+2020, Chen+2020, Sesana+2020, Shao+2021}.

@@

\lw{} is an open-source Python package designed to streamline the process of making predictions of LISA detection rates for stellar-origin binaries such that it is as fast, reliable and simple as possible. With \lw{} one can evolve the orbits of a binary or a collection of binaries and calculate their strain amplitudes for any range of frequency harmonics. One can compute the sensitivity curve for LISA or other future gravitational wave detectors (e.g. TianQin's curve, or that of a custom instrument) and use it to compute the signal-to-noise ratio of a collection of sources. Furthermore, \lw{} provides tools to visualise all of the results with easy-to-use plotting functions. Finally, \lw{} is fully tested to check for consistency in the derivations described below.

@@

We note that \lw{} is not the only initiative of this kind. We highlight the ``Gravitational Wave Universe Toolbox'' presented by \citet{Yi2021} which was developed to simulate observations on the GW universe with different detectors covering the full gravitational-wave spectrum and source classes. We also highlight ``GWPlotter'' by \citep{Moore+2015}, which provides an interactive plotting tool to compare the sensitivity of different gravitational wave detectors. \lw{} differs from the tools listed here as we have tried to provide tools that are optimised to rapidly make predictions for large populations of stellar-origin sources and we have focused on space-based detectors.

\lw{} has been developed with stellar-origin binary population studies in mind. We highlight two upcoming papers that use our package. Wagg et al.\ (in prep) investigates several populations of potential LISA sources (double black holes, black hole neutron stars and double neutron stars). They use \lw{} to evolve each synthesised source over the age of the Milky Way, make predictions about the LISA detectable population and explore how it varies with different binary physics assumptions. Additionally, Thiele et al.\ (in prep) examines the implications of assuming a metallicity-dependent binary fraction for the formation of close double white dwarfs (WDWDs) in the LISA frequency band. They use \lw{} to calculate the SNR of WDWDs and, in particular, apply a custom noise model that combines the LISA sensitivity curve from \citet{Robson+2019} with a new fit for the Galactic confusion noise based on their WDWD population.

@@

It is often convenient to convert between orbital frequency, $f_{\rm orb}$, and the semi-major axis, $a$, of a binary and this can be accomplished with Kepler's third law.
\begin{equation}{\lwFuncURL{utils}{get_a_from_f_orb}}
    a = \qty(\frac{G(m_1 + m_2)}{(2 \pi f_{\rm orb})^{2}})^{1/3},
    \label{eq:kepler3rd}
\end{equation}
where $G$ is the gravitational constant. Inversely,
\begin{equation}{\lwFuncURL{utils}{get_a_from_f_orb}}
    f_{\rm orb} = \frac{1}{2 \pi} \sqrt{\frac{G(m_1 + m_2)}{a^3}}.
    \label{eq:kepler3rd_reverse}
\end{equation}
For circular binaries, gravitational-wave emission occurs at twice the orbital frequency ($f_{\rm GW} = 2 f_{\rm orb}$). However, for eccentric binaries, we need to consider all frequency harmonics of gravitational-wave emission. These are defined such that the $n^{\rm th}$ harmonic frequency is
\begin{equation}{}
    f_{n} \equiv n \cdot f_{\rm orb}.
    \label{eq:nth_freq}
\end{equation}
It will be important to know the relative gravitational-wave power radiated into the $n^{\rm th}$ harmonic for a binary with eccentricity $e$ for the following strain and SNR calculations. This is given by \citet[][Eq.\,20]{Peters+1963}
\begin{equation}{\lwFuncURL{utils}{peters_g}}
    \begin{aligned}
        g(n, e) = \frac{n^{4}}{32} & \Big\{ \Big[ J_{n-2}(n e)-2 e J_{n-1}(n e)+\frac{2}{n} J_{n}(n e) \\
        &\left. +2 e J_{n+1}(n e)-J_{n+2}(n e)\Big]^{2}\right.\\
        &\left.\kern-\nulldelimiterspace\right.
            \!\!\begin{aligned}
            +\big(1-e^{2}\big)\Big[&J_{n-2}(n e)-2 J_{n}(n e) \\
            &+J_{n+2}(n e)\Big]^{2}
            \end{aligned}\\
        &+\frac{4}{3 n^{2}}\Big[J_{n}(n e)\Big]^{2}\Big\},
    \end{aligned}
    \label{eq:g(n,e)}
\end{equation}
where $J_{n}(v)$ is the Bessel function of the first kind. Thus, the \textit{sum} of $g(n, e)$ over all harmonics gives the factor by which the gravitational-wave emission is stronger for a binary of eccentricity $e$ over an equivalent circular binary. This enhancement factor is \citep[][Eq.\,17]{Peters+1963}
\begin{equation}{\lwFuncURL{utils}{peters_f}}
    F(e) \equiv \sum_{n = 1}^{\infty} g(n, e) = \frac{1 + \frac{73}{24} e^2 + \frac{37}{96} e^4}{(1 - e^2)^{7/2}}.
    \label{eq:eccentricity_enhancement_factor}
\end{equation}
A useful rule of thumb is that $F(0.5) \approx 5.0$, or in words, a binary with eccentricity $0.5$ loses energy to gravitational waves at 5 times the rate of an equivalent circular binary.

@@

\subsubsection{Eccentric binaries}
The orbital evolution is more complex for eccentric binaries since the semi-major axis and eccentricity both evolve simultaneously and depend on one another. The final expression cannot be solved analytically and require numerical integration. The semi major axis, $a$, and eccentricity, $e$, are related as \citep[][Eq.~5.11]{Peters+1964}
\begin{equation}{\lwFuncURL{utils}{get_a_from_ecc}}
    a(e) = c_0 \frac{e^{12/19}}{(1 - e^2)} \left(1 + \frac{121}{304} e^2\right)^{870/2299},
    \label{eq:a_from_e}
\end{equation}
where $c_0$ satisfies the initial conditions such that $a(e_0) = a_0$. The time derivative of the eccentricity, $e$, is
\begin{equation}{\lwFuncURL{evol}{de_dt}}
    \frac{\mathrm{d}e}{\mathrm{d}t} = -\frac{19}{12} \frac{\beta}{c_{0}^{4}} \frac{e^{-29 / 19}\left(1-e^{2}\right)^{3 / 2}}{\left[1+(121 / 304) e^{2}\right]^{\frac{1181}{2299}}},
    \label{eq:dedt}
\end{equation}
which we can integrate to find $e(t)$ and convert to $a(t)$ using Eq.~\ref{eq:a_from_e} (\docsLink{\lwFuncURL{evol}{evol_ecc}}).

@@

In addition, we implement the fit to Eq.~\ref{eq:t_merge_eccentric} from \citet{Mandel+2021} that approximates the merger time as
\begin{equation}{\lwFuncURL{evol}{t_merge_mandel_fit}}
    \begin{split}
        t_{\rm merge} \approx&\ t_{\rm merge, circ} (1 - e^2_0)^{7/2} \\
    &\times(1 + 0.27 e_0^{10} + 0.33 e^{20}_0 + 0.2 e_0^{1000} ),
    \end{split}
\end{equation}
which gives $t_{\rm merge}$ with an error below 3\% for eccentricities below $0.9999$. We additionally add a rudimentary polynomial fit to further reduce this error to below $0.5\%$. The user may specify whether to use this fit or perform the full integral when calculating merger times in \lw{}.

@@

\begin{figure}[htb]
    \centering
    \includegraphics[width=\columnwidth]{figures/verification_binaries_on_sc.pdf}
    \caption{The LISA verification binaries from \citet{Kupfer+2018} plotted on the LISA sensitivity curve \citep{Robson+2019} for a 4-year LISA mission. Points are coloured by their primary mass.}
    \label{fig:verification_binaries_on_sc}
\end{figure}
